---
title: "2025 Redcedar Adaptation Analysis"
output: word_document
date: "2025-03-24"
theme: cerulean
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(readxl)
```

## Introduction

Welcome to the analyses of western redcedar trees planted in for the Open Redcedar Adaptation Network 

### Research Topic

Dieback of western redcedar is increasinly reported throughout the northwest. Planting trees from southern seed sources may be a solution. The open redcedar adaptation network planted redcedar trees from two seed sources to compare the growth of trees from Oregon seed zones to trees from Washington seed zones.

### Data

Redcedar trees were planted in 2022 and remeasured each spring thereafter. The dataset analyzed here includes two years of remeasurements after planting. 

### Research Question

Is there a difference in growth rates between Oregon seed zones and Washington Seed zones. 

## Data Wrangling

```{r}
data <- read_excel("data/WRC Tree Measurements 2024.xlsx", sheet = "2024 Corrected")
```


```{r}
names(data)
```


```{r}
str(data)
```

Below we set the site and seed zone variables as factors

```{r}
data$Site <- as.factor(data$Site)
data$`Seed Zone` <- as.factor(data$`Seed Zone`)
#data$`Elk Browse` <- as.factor(data$`Elk Browse`)
```





## Data Visualization


Lets look at the distribution of the data


```{r}
ggplot(data, aes(`Tree Height (cm)`) ) +geom_histogram() +theme_bw()
```

It looks like there are a lot of trees in the 50cm range. Perhaps that is related to elk browse?

Let try to plot it without any of the trees that had elk browse

```{r}
ggplot(data,aes(`Elk Browse`))+geom_bar() +coord_flip()+theme_bw()
```

```{r}
data.browsed <- data %>% mutate(`Elk Browse` = recode(`Elk Browse`, YEes = 'Yes', Low = 'Yes', High =  'Yes' ))
```




```{r}
str(data.browsed)
```


```{r}
data.browsed$`Elk Browse`[is.na(data.browsed$`Elk Browse`)] <- "NA"
```


```{r}
ggplot(data.browsed,aes(`Elk Browse`))+geom_bar() +coord_flip()+theme_bw()
```


```{r}
data.browsed.filtered <- data.browsed %>% filter(`Elk Browse` != "Yes")
```



```{r}
ggplot(data.browsed.filtered, aes(`Tree Height (cm)`) ) +geom_histogram() +theme_bw()
```


```{r}
ggplot(data.browsed.filtered,aes(`Seed Zone`)) +geom_bar() +theme_bw() 
```


Oh, we have one tree in seed zone labeled 'Wa' instead of 'WA'

```{r}
data.browsed.filtered$`Seed Zone`[data.browsed.filtered$`Seed Zone`=="Wa"] <- "WA"
```

```{r}
ggplot(data.browsed.filtered,aes(`Seed Zone`)) +geom_bar() +theme_bw()
```


```{r}
ggplot(data.browsed.filtered,aes(`Seed Zone`,fill=Site)) +geom_bar() +theme_bw()
```


Lets filter out the Puyallup site

```{r}
data.browsed.site.filtered <- data.browsed.filtered %>% filter(Site!="Puyallup")
```



```{r}
ggplot(data.browsed.site.filtered, aes(`Tree Height (cm)`) ) +geom_histogram() +theme_bw()
```


```{r}
ggplot(data.browsed.filtered,aes(`Seed Zone`,`Tree Height (cm)`)) +geom_boxplot() +theme_bw()
```


```{r}
#ggplot(df, aes(x=Category, y=Mean, fill=Quality)) +  geom_bar(position=position_dodge(), stat="identity",  colour='black') + geom_errorbar(aes(ymin=Mean-sd, ymax=Mean+sd), width=.2)
```



## Analyses 

Lets compare the standard deviation between groups

```{r}
library(car)
leveneTest(`Tree Height (cm)` ~ `Seed Zone`, data=data.browsed.site.filtered)
```

According to the Levene's Test for Homogenity of Variance, the standard deviation in tree height between Oregon and Washington seed zones is significantly different. reject the null -

"tests the null hypothesis that the population variances are equal (called homogeneity of variance or homoscedasticity). If the resulting p-value of Levene's test is less than some significance level (typically 0.05), the obtained differences in sample variances are unlikely to have occurred based on random sampling from a population with equal variances. Thus, the null hypothesis of equal variances is rejected and it is concluded that there is a difference between the variances in the population." - Wikipedia article https://en.wikipedia.org/wiki/Levene%27s_test

Basically, there is a difference in variation. Therefore, an anova can't be used to compare the tree heights in the data. 

Determining the tests that are appropriate

Depending on the distribution and shape of the data, we may or may not be able to compare groups with anova or t-tests. 



```{r}
ggplot(data.browsed.filtered,aes(`Tree Height (cm)`,`Tree Diameter (mm)`, fill=`Seed Zone`)) +geom_smooth()
```




```{r}
model.one <- glm(`Tree Height (cm)` ~ `Seed Zone` + Site, data=data.browsed.site.filtered)

summary(model.one)
plot(model.one)
```






## Results 

